---
title: "Logistic Regression Analysis Example: Optimizing Expected Returns"
date: "2026-01-13"
description: "A simulated study on how to build a logistic regression model and utilize it to optimize a target variable"
categories:
  - "Machine Learning"
  - "Logistic Regression"
  - "Optimization"
engine: knitr
freeze: auto
toc: true
toc-depth: 2
toc-location: left
execute:
  results: hold
---

Last Update: `r format(Sys.Date(),"%B %d, %Y")`

```{r}
#| warning: false
#| message: false
library(knitr)
library(kableExtra)
library(arm)
library(gtsummary)
library(dplyr)
library(ggplot2)
library(future.apply)
```

# Introduction

This will be the one of the first in a series of example analyses I plan to post online to showcase my statistical technical knowledge as well as document some interesting problems I have come across in my experience. I specifically want to focus on what and why I do things over the course of an analysis rather than the how I do the analysis. As such, I will assume some familiarity with the statistical tools and code presented here and generally keep theoretical discussion at a higher level.

The article is organized into the following parts:

1.  A basic introduction of the logistic regression model and its comparison to other machine learning models

2.  A simulated example detailing the workflow to build a logistic regression model [jump to section](#prob)

3.  Methods to leverage the logistic regression model to optimize some target goal [jump to section](#sec-optimize)

Readers who are already familiar with logistic regression models may want to skip to the section that details the analysis question and then skip to the section on how to optimize for a target goal.

# Logistic Regression

Logistic regression is one of the fundamental tools of machine learning and statistics and is actually one of the more powerful tools when one has enough time to go that extra mile and fine tune the model to the data. First I'll just state the mathematical formula for logistic regression beginning with the data model:

$Y_i|\pi_i \sim Bernoulli(\pi_i)$

where $i$ refers to the data sample index (think of a row in your data table), $Y_i$ is whether "success" is observed for sample $i$, and $\pi_i$ is the probability of success for the $i$-th subject. In typical applications, $Y_i$ is coded such that 0 means "failure" and 1 means "success". Now we can move on to the regression part of the model between the log-odds and the covariates of interest:

$\log(\frac{\pi_i}{1-\pi_i}) = \pmb x_i \pmb \beta$

where $\pmb x_i$ is a vector of covariates that you think explain the probability of success and $\pmb \beta$ is the vector of weights that indicate how strongly they contribute to the log-odds of success. Vectors are simply shorthand notation to represent collection of all the covariates in the model and their corresponding parameters.

When estimating the logistic regression model, the $\pmb \beta$ coefficients are the main parameters of interest (as $\pi_i$ is derived from $\pmb \beta$). The reason we call it "logistic" regression is because $\log(\frac{x_i}{1-x_i})$ is the inverse of the logistic function.

The assumptions of the model are as follows:

1.  The outcome is a binary random variable, specifically a Bernoulli[^1] random variable.

2.  The independent/explanatory variables (I'll commonly refer to these as covariates) have a linear relationship with the log-odds of "success". The linearity here refers to the beta parameters, for example one may include an $x^2\beta$ term in the model as the beta parameter has a power of 1.

3.  Observations are independent of each other

4.  No multicollinearity between covariates. For any single covariate, the other covariates in the model don't perfectly describe that covariate.

5.  No complete separation. In other words, the covariates cannot perfectly predict binary outcome.

[^1]: There is an extension of logistic regression to group level data where the outcome is the number of successes observed and is modeled as a binomial random variable.

## Logistic regression compared to other machine learning models

The important distinction between logistic regression and other machine learning models is that the statistical theory on the model parameters is well understood. For each of the $\beta$ parameters in our model, we know the statistical properties to allow inference through p-values or confidence intervals. Logistic regression is still a tool used today to understand, or *infer*, the relationship between a binary outcome and other data collected when statistical evidence is required to validate any findings.

Although in most real-world scenarios the other machine learning models perform better in terms of predictiveness, logistic regression can sometimes still be a better model in terms of predictiveness. In my experience through classes and work projects, a logistic regression model is actually superior to the more sophisticated machine learning techniques of Gradient Boosted Trees and Random Forests under a few scenarios with respect to the AUC (area under the curve) performance metric.

The first scenario I've come across is when the sample size is relatively small. The machine learning techniques tend to work very well when there is more data to help effectively learn the functional relationship between the data you have and the target outcome. I would say this would be the situation when your sample size is in the magnitude of a few hundreds or less. After that point, other machine learning techniques perform at their expected capacity.

The second scenario where logistic regression can prove better than tree based methods is if one is willing to both select the covariates that strongly relate to the outcome of interest and also identify how each covariate functionally relates to the outcome. To expand on what I mean by "how each covariate functionally relates to the outcome", some examples of functional relationships are the usual linear relationship, quadratic relationship, or sinusoidal relationship. In mathematical terms you can propose the following models:

$Linear: \log(\frac{\pi_i}{1 - \pi_i}) = x_i\beta$

$Quadratic: \log(\frac{\pi_i}{1 - \pi_i}) = x_i\beta_1 + x_i^2\beta_2$

$Sinusoidal: \log(\frac{\pi_i}{1 - \pi_i}) = sin(x_i)\beta$

Choosing a functional relationship alone produces infinite possibilities for one to sort through in addition to just choosing which covariates to include in one's analysis. The strength of tree based approaches is that they learn these relationships with enough data, but of course the tree won't be as accurate as if one were to explicitly identify that "true" relationship when training a logistic regression. However, identifying the "true" functional relationship between the outcome and covariates is rather impractical when one has many different covariates to consider at once. So once again, when the data becomes sufficiently large, it is more practical to use a Gradient Boosted Tree model or Random Forest model.

The advantages and shortfalls of logistic regression compared to other machine learning techniques naturally extend to any generalized linear model such as linear regression. To be clear, I am **NOT** saying that in these scenarios generalized linear models will **ALWAYS** surpass the newer machine learning models and algorithms, but rather that these scenarios **ALLOW** generalized linear models to shine just as well, if not better, than their other machine learning counterparts in terms of predictiveness.

# A Novel Application Problem {#prob}

Suppose that one is starting an online clothing store selling clothes, shoes, and accessories. They have the production cost of each product and wish to know what markup percentage is optimal for generating the most revenue. The data is set up such that we have many offers for products on sale for different markup percentages and it is recorded whether the sale was made or not. From past data, it is plausible that the type of product (clothes, shoes, accessories) has an affect on whether a sale is made.

Defining and estimating the optimal markup percentage is the novel part of the analysis that I think differentiates this analysis from typical logistic regression analyses where the focus is usually on determining a statistically significant relationship between a covariate and the outcome of interest.

# Simulating Toy Data {#sec-simdata}

Before proceeding with the analysis, we shall simulate a toy data set in R that will fit the problem. In the ideal case, I would have found a real world data set that matches our example, but my superficial search of the internet returned only aggregated product sales and wouldn't quite match the situation described above. This does detract from exploratory analyses as I know how the data is generated and greatly simplifies the model building process. However, the general workflow and considerations for building the model would be the same with a real data set.

We will use the following logistic regression model to simulate data:

$$
\log(\frac{\pi_i}{1 - \pi_i}) = \beta_0 + x_{markup} * \beta_{markup} + 
x_{cost} * \beta_{cost} + x_{type=shoes} * \beta_{type=shoes} + x_{type = clothes}*\beta_{type = clothes}
$$

where $x_{markup}$ is the markup percentage above 0 (above 100 is allowed), $x_{cost}$ is cost in dollars, $x_{type=shoes}$ is 1 when product $i$ is shoes and 0 otherwise, and similarly $x_{type = clothes}$ is 1 when product $i$ is clothes and 0 otherwise. When product $i$ is accessories, both $x_{type=shoes}$ and $x_{type = clothes}$ are 0. The way that $x_{type}$ is encoded is known as the dummy variable encoding common in statistics (which is different from the one-hot encoding more prevalent in the machine learning field). The outcome of interest $Y_i$ is a binary variable where 1 indicates the product was sold and 0 the product was not sold, and it follows that $\pi_i$ is the probability that product $i$ is sold or not. To clarify further about the $x_{markup}$ covariate, if $x_{markup} = 20$ then the product was marked up $20\%$.

I chose the values for true parameters $\pmb \beta$ that I felt gave a good spread of the true probabilities while also being representative of the expected behavior in the example problem described to me. I arbitrarily chose a decently large sample size of 3000 to ensure proper estimation.

```{r}
## True parameter values

set.seed(2013)

sample_size <- 3000

true_beta <- c(2, # intercept
               -0.2, # markup
               0.002, # cost
               1, # type=clothes
               1.5 )# type=shoes

### Simulate data

type <- sample(c("accessories", "clothes", "shoes"), 
               size = sample_size,
               replace = TRUE,
               prob = c(0.3, 0.5, 0.2))
cost <- rnorm(sample_size, mean = 2000, sd = 400)
markup <- runif(sample_size, min = 20, max = 120)

## make sure markup and cost are above 0
markup <- ifelse(markup <= 0, 10, markup)
cost <- ifelse(cost <= 0, 2000, cost)

# convert type into a factor variable to prepare for analysis

type <- factor(type, levels = c("accessories", "clothes", "shoes"))

x_df <- data.frame(markup = markup, cost = cost, type = type)

# get x_matrix

x_matrix <- model.matrix(~ 1 + markup + cost + type, data = x_df)

#simulated log-odds per person

log_odds <- x_matrix %*% true_beta

#convert log_odds to probabilities

inverse_logit <- function(x){(1 + exp(-x))^-1}
true_probs <- inverse_logit(log_odds)

# get the outcome
sold <- rbinom(sample_size, 1, true_probs)
x_df$sold <- sold
# x_df$true_probs <- c(true_probs)
# x_df$true_log_odds <- c(log_odds)
```

Here are the first few rows of the data set to get an idea of how the data is organized and its values:

```{r}
head(x_df, 5) %>%
  kable(align = 'c') %>%
  kable_styling()
```

# Example Logistic Regression Analysis

Before jumping into the analysis or looking at the data, it is imperative to determine whether the analysis is attempting to optimize prediction or inference. Although a model that infers relationships well typically predicts well and vice versa, the criteria statisticians use to judge what model is best typically favor one over the other. Because of the optimization scheme I outline in a [later section](#sec-optimize), I believe having an unbiased estimate of $\pi_i$ and $\beta_{markup}$ is the main goal, so I interpret this as predominantly an inference problem because of that.

## Quick Exploratory Analysis of the Data Set

When starting an analysis, it's important to first summarize and get a feel of the data. This is a useful and important step as it informs what are reasonable models for the data and indicates the quality of the data. The `gtsummary` package is used to produce the pretty summary tables presented below.

```{r}
#| label: tbl-summ
#| tbl-cap: "Summary Tables"
#| tbl-subcap: 
#|   - "Summary by Type"
#|   - "Summary by Sold"
#| layout-nrow: 2
x_df %>% 
  tbl_summary(
    include = c(markup, cost, type, sold),
    type = list(all_continuous() ~ "continuous2"),
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})", "[{min}, {max}]")
    ),
    by = type)

x_df %>%
  tbl_summary(
    include = c(markup, cost, sold),
    type = list(all_continuous() ~ "continuous2"),
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})", "[{min}, {max}]")
    ),
    by = sold)
```

```{r}
#| label: fig-comp
#| fig-cap: "Comparing cost and markup"

plot(markup, cost, 
     main = "cost vs markup", 
     ylab = "cost of production ($)", 
     xlab = "markup of product (%)")
```

The first table summarizes the different covariates and outcome by type. Inspecting the average and quartiles of markup and cost between the different types, we don't notice any difference in the distributions. This implies that type is associated with neither the cost nor markup, and this is a best case scenario for a logistic regression analysis. The proportion of successful sales is different among the product types with clothes and shoe products selling at a higher rate than the accessories. This implies that there may be an association between type and product sales. Clearly, there is an imbalance in the distribution of type in the data set, but we have enough cases of each product type such that including type in the logistic regression model shouldn't be a problem.

The second table summarizes markup and the cost variables by whether the product sold (the 1 group) or not (the 0 group). Comparing these two groups, we can clearly see that markup is lower in the groups that sold and indicates a negative association between sales and markup. The distribution of cost shifts about 100 dollars higher in the group that sold and thus suggests a positive association between cost and sales.

Finally, in the scatter plot comparing cost and markup, we observe random spread of the points. There is no indication of a linear or other functional relationship between markup and cost, which is an ideal scenario for logistic regression.

## Building a logistic regression model

Fitting a logistic regression model in R is a simple one line of code. Since we know what the true generating parameters are above, let's double check our work by comparing the fitted values to the true parameters.

```{r}
#| label: tbl-truefit
#| tbl-cap: "Comparing estimates to true values. Estimates are on the log-odds scale"
fit <- glm(sold ~ 1 + markup + cost + type, 
           data = x_df,
           family = binomial)
ci <- confint(fit)
result <- data.frame(truth = true_beta, 
                     est = coef(fit),
                     ci_lwr = ci[,1],
                     ci_upr = ci[,2])

kable(result, digits = 4, align = 'c') %>% 
  kable_styling()
```

As expected with this rather large sample size, the estimates are fairly close to the true values and are well within the 95% confidence intervals. All of the parameter estimates are statistically significant as 0 is not within the confidence intervals, so this is a convincing model. At this step, one would normally interpret these values as odds ratios, but that is not the goal of this analysis, so I will skip the interpretation.

When building a logistic regression model, one usually has to consider a few different models to find the best fit to the data. Let's consider three other logistic regression models. One without type in the model but all other variables in the model:

$$
\log(\frac{\pi_i}{1 - \pi_i}) = \beta_0 + x_{markup} * \beta_{markup} + 
x_{cost} * \beta_{cost}
$$

a second model with a quadratic term for markup and including all the other variables:

$$
\log(\frac{\pi_i}{1 - \pi_i}) = \beta_0 + x_{markup} * \beta_{markup} + x_{markup}^2 * \beta_{markup^2} +
x_{cost} * \beta_{cost} + x_{type=shoes} * \beta_{type=shoes} + x_{type = clothes}*\beta_{type = clothes}
$$

and finally a simple model with just markup:

$$
\log(\frac{\pi_i}{1 - \pi_i}) = \beta_0 + x_{markup} * \beta_{markup}
$$

There are many model selection criteria to choose from and which one to use depends on the goal of the analysis. In this application, the goal is estimation and inference, so I would personally advocate for the Bayesian Information Criterion (BIC). The BIC is a quantity that expresses how well the proposed model fits the data while also balancing how many parameters are in a model. More parameters in a model results in a better fit to the data but too many parameters results in overfitting. The problem with an overfit model is the lack of generalizability to new data. Please refer to the [BIC() documentation](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/AIC) to see the formula for its calculation. BIC favors having less parameters in a model as compared to it's counterpart the Akaike Information Criterion (AIC). Therefore, the BIC is sometimes the preferred model selection criterion when the goal is to identify what covariates to go into a model.

Lower values of BIC indicate a better statistical model for the data. A difference of at least 2 is the typical threshold to consider one model superior the other. For differences of less than 2 the models are about equivalent and the statistician must decide which model is superior.

```{r}
#| label: tbl-modselect
#| tbl-cap: "Comparing Logistic Regression Models"
fit1 <- glm(sold ~ 1 + markup + cost, 
           data = x_df,
           family = binomial)
fit2 <- glm(sold ~ 1 + markup + I(markup^2) + cost + type, 
           data = x_df,
           family = binomial)
fit3 <- glm(sold ~ markup, data = x_df, family = binomial)

df_model_selection <- data.frame(model = c("True_model", "no_type", "quadratic_markup", "Simple_markup"), 
                                 bic = c(BIC(fit), BIC(fit1), BIC(fit2), BIC(fit3))
                                 ) 
df_model_selection %>%    
  kable(digits = 2, align = 'c', caption = "Smaller BIC indicates better model") %>%  
  kable_styling()

```

Upon inspecting the BIC values, the true model is unsurprisingly the lowest BIC and thus the best model. Disregarding our knowledge of the true data generating model, the model chosen at this step strongly translates the trends observed in the exploratory data analysis step, and thus highlights the importance of exploring the data before building models. The simple model with just markup is clearly the worst model with the highest BIC, so although the simple model would drastically simplify our optimal estimation of $\pi_i$ in the next step, the data indicates a more complex relationship between the outcome and covariates is more appropriate.

Although we only inspect 4 models here, usually one would explore many more models and spend the bulk of their time trying to figure out the best model. Usually background knowledge of important factors is involved when building the model and some consideration of limitations in the data quality. More complex relationships like interactions and different functional transformations of continuous variables could also be considered. Based on the exploratory analysis, this does not seem necessary in this analysis.

## Assessing Model Assumptions

Assessing whether the model assumptions are violated is often overlooked when fitting a logistic regression model. The main ones to assess are linearity of covariates (assumption 2) and multicollinearity (assumption 4). The independent sampling assumption (assumption 3) is usually considered when planning the analysis and can potentially investigated through exploratory analysis. Things to look for include covariates that strongly correlate, categorical values that do not have any observed success, and background knowledge about whether independence makes sense. Perfect separation (assumption 5) is normally a warning from R if/when that occurs.

### Assessing Multicollinearity

Based on the exploratory analysis earlier, multicollinearity is not likely not an issue in this data set. We know this for a fact because I generated the data independently. The exploratory analysis revealed that there wasn't any clear relationship between pairs of the covariates, and all of the parameters in the model are statistically significant. These are both great signs that multicollinearity is likely not present. However, to formally check for multicollinearity, one can check the Variance Inflation Factors (VIFs) in the covariates. I'll omit checking the VIFs due to the exploratory analysis revealing a high quality data set, but one can read more on VIF theory here: [article](https://online.stat.psu.edu/stat462/node/180/) and how to calculate it using the `car` package in R here: [article](https://www.r-bloggers.com/2023/12/exploring-variance-inflation-factor-vif-in-r-a-practical-guide/).

I will note though that if one adds higher order terms in the regression, for example,

$$
\log(\frac{\pi_i}{1 - \pi_i}) = \beta_0 + x_{markup} * \beta_{markup} + x_{markup}^2 * \beta_{markup^2}
$$ 
then instead of the VIF one will need to inspect the Generalized Variance Inflation Factor (GVIF).

### Assessing linearity

The hardest assumption to verify and is most often overlooked is the linearity assumption. This only applies to continuous covariates which are markup and cost in this example. We can use the `binnedplot` function in the `arm` package to assess this. What this function does is bin or discretize the continuous variable into equally spaced segments. Within each segment, calculate the residual for the observations via:

$$
  residual_i = y_i - \hat \pi_i 
$$ 

where $\hat \pi_i$ is calculated from a proposed logistic regression model. We inspect the plot to make sure that the residuals are randomly scattered around the $y = 0$ horizontal line, there is no pattern in the residuals, and that at most \~5% of the residuals are outside of the confidence interval lines.

Below is a plot to illustrate how the binned residuals are calculated. The points plotted in a binned residual plot would be the points observed in red.

```{r}
## example on how the binned residual plot is constructed
rng <- range(x_df$cost) # min, max of markup
## break up range into 3 equal parts
breakpoints <- seq(from = rng[1],
                   to = rng[2],
                   by = diff(rng/3))
midpoints <- (breakpoints[-1] + breakpoints[-length(breakpoints)]) / 2
res <- x_df$sold - predict(fit, type = 'response')
res_df <- data.frame(res = res,
           bp = cut(x_df$cost, breaks = breakpoints))
res_ave <- aggregate(res ~ bp,
                     data = res_df,
                     FUN = mean)
plot(x = x_df$cost, y = res, 
     xlab = "cost",
     ylab = "Residuals: observed - predicted",
     main  = "How a binned residual plot is constructed",
     pch = 16)
points(x = midpoints, y = res_ave$res, 
       pch = 16, cex = 1.2,
       col = 'red')
abline(v = breakpoints, col = 'red')
legend("topright", 
       legend = c("observed_residual",
                  "average_residual"),
       col = c('black', 'red'),
       pch = 16)
```

These binned residuals plots assume a large sample size for the statistical properties to hold. Specifically **if the sample size is large within each bin**, then the expected value of these residuals is 0 with known standard errors due to the Lindeberg-Feller central limit theorem. My personal rule-of-thumb is at least 30 observations to be reasonably confident in the approximation, but the more the better. Unfortunately, the binned plots from the arm package don't throw warnings when the bins have small sample size, so some caution is needed when interpreting these plots. In addition to large sample sizes, the statistical theory does not hold for values of $\hat \pi_i$ close to 0 or 1, so bins with predicted values of $\hat \pi_i$ close to 0 or 1 are not as informative.


Let's look at several binned residual plots to assess whether our proposed model is reasonable. The first plot below is binned according the the predicted $\pi_i$ values. This is a nice overall assessment in the model and usually a good starting point of whether assumptions are violated. Most of the binned data points are within the theoretical confidence intervals and fairly close to 0. There isn't a strong pattern that I discern from the residuals, so I am personally satisfied with this plot.

```{r}
res <- residuals(fit, type = "response")
binnedplot(x = fitted(fit),
           y = res, 
           nclass=NULL, 
           xlab="Expected Values", 
           ylab="Average residual", 
           main="Binned residual plot")
```

The next plot below is a binned residual plot where the bins are made according to the markup covariate. The plot seems fine nice and random and within the confidence intervals until the point where markup is 60\% or higher where the binned plot degenerates to 0. This is an expected pattern due to no sales being made for markups above 57\%,

```{r}
binnedplot(x = x_df$markup,
           y = res, 
           nclass=NULL, 
           xlab="Binned markup", 
           ylab="Average residual", 
           main="Binned residual plot of markup")
```

Below is a binned plot where the bins are defined by the cost variable. This is a nice plot where the random scatter is fairly tight and the values are almost all within the confidence intervals.

```{r}
binnedplot(x = x_df$cost,
           y = res, 
           nclass=NULL, 
           xlab="Binned cost", 
           ylab="Average residual", 
           main="Binned residual plot of cost")
```

Based on the binned plots above, the logistic regression model strongly satisfies the linearity assumption. If there were any patterns in the residual plots, then one would need to refine the logistic regression model with higher order terms (i.e. $x_{cost}^2$) or nonlinear transformations (i.e. $\log(x_{cost})$) to address the patterns. This is not necessary by design, but a future post of mine will explore how to spot different patterns in residual plots.

# Optimizing the Expected Returns (Expected Revenue) {#sec-optimize}

Now that we have built a satisfactory logistic regression model, we may now proceed to calculating the optimal markup percentage. First we will show how to estimate $\pi_i$ using the logistic regression model and how we can calculate the optimal markup percentage.

## Relationship between $x_{markup}$ and $\pi_i$

Now that we have decided on a logistic regression model, let's explore the relationship between markup and the probability of selling the product. The relationship between markup and the log-odds might be linear, but when we invert the problem to get the relationship between markup and the probability it is certainly non-linear. To produce this non-linear relationship, we can invert the log-odds back to the probability scale through the following formula:

$$
\pi_i = (1 + \exp[-(\beta_0 + x_{markup} * \beta_{markup} + 
x_{cost} * \beta_{cost} + x_{type=shoes} * \beta_{type=shoes} + x_{type = clothes}*\beta_{type = clothes})])^{-1}
$$

We are specifically interested in how markup relates to the probability of selling, but we have the other data variables in the formula. Before exploring how the probability of selling changes over a range of values for markup, it is pertinent to choose some representative values for cost and type. To start with, let's set the cost level to be the observed average cost level and let type be the most common type in the data set to provide a curve.

```{r}
## there isn't a mode function in base R
getMode <- function(x){
  lx <- levels(x)
  ux <- unique(lx)
  factor(ux[which.max(tabulate(match(lx, ux)))], levels = lx)
}
range_markup <- seq(from = 0, to = 75, by = 1)
mode_type <- getMode(x_df$type)
mean_cost <- mean(x_df$cost)
new_x_df <- data.frame("markup" = range_markup,
                       "cost" = mean_cost,
                       "type" = mode_type)

preds <- predict(fit, newdata = new_x_df, se.fit = TRUE)
preds_ci_upr <- preds$fit + qnorm(0.975)*preds$se.fit
preds_ci_lwr <- preds$fit - qnorm(0.975)*preds$se.fit
probs_df <- data.frame(pred_prob = inverse_logit(preds$fit),
                       pred_prob_lwr = inverse_logit(preds_ci_lwr),
                       pred_prob_upr = inverse_logit(preds_ci_upr),
                       markup = range_markup)

true_probs <- model.matrix(~ markup + cost + type, data = new_x_df) %*% true_beta
probs_df$true_probs <- inverse_logit(true_probs)


ggplot(probs_df) + 
  ## probability curve predicted from logistic regression in black
  geom_line(aes(x = markup, y = pred_prob)) + 
  ## 95% wald CI in red
  geom_line(aes(x = markup, y = pred_prob_upr), col = "red") + 
  geom_line(aes(x = markup, y = pred_prob_lwr), col = "red") + 
  ## probability curve using true values will be in blue
  geom_line(aes(x = markup, y = true_probs), col = "blue") + 
  ggtitle(paste0("Probablity of Selling Product vs markup percentage"),
          subtitle = paste0("type = ", mode_type[1], 
                            "  cost = ", 
                            round(mean_cost[1], digits = 2)
                            )
  ) + 
  labs(x = "markup (%)", y = "Probability of Selling Product")

```

As expected, due to how well our estimates of the parameters are, the estimated probability curve in black is mostly overlapped by expected probabilities calculated from the true parameters in blue. As we can see from the graph above, the probability of selling decreases as markup increases. This isn't surprising given that the true parameter for markup is -0.02.

Let's plot the differences between the true and predicted probabilities to really identify the small the difference in magnitude. We can see that the probabilities only differ at most by 0.02. Depending on the application context, this can be a meaningful difference or not, but for this application let's assume that an underestimation of 0.02 is a reasonably close estimate.

```{r}
#| label: fig-diffprob
#| fig-cap: "Differences between predicted probability and true probability"
plot(x = probs_df$markup, 
     y = probs_df$true_probs - probs_df$pred_prob,
     type = 'h',
     ylab = "True Prob - Pred Prob",
     xlab = "markup",
     main = "Difference between True and Predicted Probabilities")
```

## Calculating the Optimal Markup

We can propose the following function to express a balance between optimizing cost markup while increasing the probability of successfully selling the product:

$$
f(x_{markup}) = (1 + \frac{x_{markup}}{100})x_{cost} \pi_i
$$

This function is derived from the idea of the expected (aka average) total revenue generated by a new product. The new products total revenue is expressed as $(1 + \frac{x_{markup}}{100})*x_{cost}$ and the expected value of a new observation $Y_i$ is $\pi_i$ because $Y_i$ is a Bernoulli random variable (see [appendix](#sec-exp-return) for a small statistical derivation of this quantity). Thus, the expected total revenue generated from this product is the multiplication of these two quantities.

In the previous section, we calculated how to estimate $\pi_i$ using the logistic regression model and some chosen values for cost and type. So we have all of the needed components to calculate the function above and now we need to determine two things:

  1. Whether a maximum value actually exists

  2. What value of $x_markup$ maximizes the function.
  
The first part is rather easy to show with a graph below. Cost is set to the observed average cost in the data set (`r mean(x_df$cost)`) and type is set to the most common type of product in the data set `r getMode(x_df$type)`.

```{r}
expected_return <- function(markup, cost, type, fit, log = FALSE){
  new_data <- data.frame(markup = markup,
                         cost = cost,
                         type = type)
  
  if(log){
   ### when doing optimization, the log scale is more stable if there are convergence issues
   ### need to restrict if markup > 0
    log(1 + markup/100) + log(cost) + 
      log(predict(fit, newdata = new_data, type = "response"))
  }else{
   (1 + markup/100)*cost * #total sale price
    predict(fit, newdata = new_data, type = "response") #prob success 
  }
}

mode_type <- getMode(x_df$type)
mean_cost <- mean(x_df$cost)
x_markup_vals <- seq(0, 120, by = 0.1)
exp_return_vals <- sapply(x_markup_vals, 
                          expected_return, 
                          cost = mean_cost, 
                          type = mode_type, 
                          fit = fit,
                          log = FALSE)

plot(x_markup_vals, 
     exp_return_vals, 
     type = 'l',
     xlab = "markup (%)",
     ylab = "Expected Return (Revenue)",
     main = "Expected Revenue vs Markup Percentage")
```

From the graph above, it is clear that a maximum occurs around where markup is ~20% and the expected return is about ~2100. To calculate the exact value for the optimal markup and expected return, we will need to use a numerical optimization algorithm. In R, we can accomplish this using the `optim()` function. There are several different algorithms to choose from, in this case Broyden–Fletcher–Goldfarb–Shanno (BFGS) quasi-newton method is a great choice. Going into detail of the BFGS algorithm is beyond the scope of this project. The [wikipedia page](https://en.wikipedia.org/wiki/Broyden%E2%80%93Fletcher%E2%80%93Goldfarb%E2%80%93Shanno_algorithm) is an excellent starting point for more information. 

Important outputs to note from the `optim()` function is `par` which is the markup value where the maximum occurs, `value` which is maximum revenue value, and `convergence` which should be 0 so that you know that the algorithm completed successfully.



```{r}
optimal_vals <- optim(
      # initical guess
      par = 20, 
      # function to optimize
      fn = expected_return, 
      # optimization algorithm
      method = "BFGS",
      # do maximization, default is minimization
      control = list(fnscale = -1),
      # Additional params for function
      cost = mean_cost[1], 
      type = mode_type[1], 
      fit = fit 
      )


if(optimal_vals$convergence != 0){
  stop("The numerical optimization did not converge")
}else{
  new_data <- data.frame(markup = optimal_vals$par,
                         cost = mean_cost[1],
                         type = mode_type[1])
  optimal_prob_success <- inverse_logit(predict(fit, new_data))
  optimal_df <- data.frame(optimal_markup = optimal_vals$par,
                           expected_total_cost = optimal_vals$value,
                           type = mode_type[1],
                           cost = mean_cost[1],
                           prob_success = optimal_prob_success)
  kable(optimal_df, digits = 3, align = 'c') %>%
    kable_styling()
}
```

From this analysis, we find that for an accessory product that costs of \$`r round(optimal_df$cost, digits = 2)` to produce, the optimal markup is about `r round(optimal_df$optimal_markup, digits = 2)` percent. This results in a total revenue generated of \$`r round(optimal_df$expected_total_cost, digits = 2)`.

Although our formula calculates an optimal markup of `r round(optimal_df$optimal_markup, digits = 2)` percent, we need to keep in mind that the range of markup in the data set are values between 20 and 120 percent. Thus, a markup of `r round(optimal_df$optimal_markup, digits = 2)` percent is outside the region covered by the data. I would suggest collecting more data in this range before widely adopting this decision rule.

In this section we only looked at one specific combination of covariates where the product cost is equal to the average cost and the product type is the most common type in the data set. Of course, we may want to consider cases with combinations of product costs and product types. To explore these different scenarios more easily, I've created a Rshiny dashboard that'll be linked below when it's done.

## Hypothetical Impact of Optimal Markup strategy

Let's assess the business impact of this optimal markup strategy with regards two important business performance metrics:

1.  The change in successful sales compared to the observed sales data

2.  The change in revenue generated compared to the observed revenue

For each point in our data set, we will substitute the observed markup value with the optimal markup value and simulate whether the sale is successful or not. To account for extreme cases, such as when a 99\% chance sale fails, I will replicate the experiment 5000 times and calculate the average successful sales and revenue generated and compare these values to the observed revenue and sales.

Calculating the optimal markup for the entire data set is a bit slow, so I used the `future.apply` package in R to run the the calculation in parallel. Going into detail about the `future.apply` package and the `future` backend is beyond the scope of this project, but I'll make a blog post about it sometime in the future (pun not intended).

```{r}
calc_optimal_markup <- function(cost, type, fit){
  optim(par = 20, 
      fn = expected_return, 
      method = "BFGS",
      control = list(fnscale = -1), # do maximization not minimization
      # These params don't change
      cost = cost, type = type, fit = fit 
      )  
}

# non-parallel (sequential) version

# optimal_markups <- mapply(calc_optimal_markup,
#                           cost = x_df$cost,
#                           type = x_df$type,
#                           MoreArgs = list(fit = fit)
#                           )


# set up parallel computing
plan("multisession")
# do the calculation for every data point in parallel
# on my machine this speeds up the calculation about 6 fold
optimal_markups <- future_mapply(calc_optimal_markup, 
                                 cost = x_df$cost, 
                                 type = x_df$type, 
                                 MoreArgs = list(fit = fit)
                                 )
# return computing environment back to normal
plan("sequential")

optimal_x_df <- x_df
optimal_x_df$markup <- unlist(optimal_markups[1,])

## predicted probabilities with optimal markup
hypothetical_probs <- predict(fit, 
                              newdata = optimal_x_df, 
                              type = 'response')

### simulate whether a sale was made for the optimal markup data set, 5000 times
set.seed(2013)
expected_results <- replicate(5000,
          expr = {
            hypothetical_success <- rbinom(sample_size, 1, hypothetical_probs)
            c(change_sales = sum(hypothetical_success) - sum(x_df$sold),
              change_rev = sum(hypothetical_success * (1 + optimal_x_df$markup/100)*x_df$cost) - 
                           sum(x_df$sold * (1 + x_df$markup/100)*x_df$cost)
            )
          })
```


```{r}
## create table of observed and optimal results
obs_revenue <- sum(x_df$sold * (1 + x_df$markup/100)*x_df$cost)
optimal_results <- data.frame(observed = c(sum(x_df$sold), obs_revenue),
                              sim_ave = c(mean(expected_results[1,]),
                                          mean(expected_results[2,]))
                              )

optimal_results <- optimal_results %>%
                    mutate(percent_change = 
                             100*ifelse(sim_ave/observed > 1, 
                                        sim_ave/observed - 1, 
                                        1 - sim_ave/observed),
                           difference = sim_ave - observed
                          )
rownames(optimal_results) <- c("Sales", "Revenue")

## format the results for easier reading
optimal_results <- as.data.frame(t(optimal_results))
optimal_results %>% mutate(revenue = revenue)
# optimal_results[2, c(1, 2, 4)] <- 
label_currency(unlist(optimal_results[2, c(1, 2, 4)]))
rownames(optimal_results) <- c("Sales", "Revenue")

## print results
optimal_results %>% 
  kable(digits = 2, align = 'c') %>%
  kable_styling() %>%
  footnote("Revenue per 1000")
```

Write up the results

# Discussion and Conclusion

Easy to extend to other calculations that take into account profit margins such as tax calculations

Can identify underperforming/unfeasible products

Does not consider external economical factors or 

# Appendix 1: Deriving Expected Revenue Formula {#sec-exp-return .appendix}

First, we calculate that $(1 + \frac{x_{markup}}{100})x_{cost}$ would be the sales price of the item (ignoring any taxes).

Next let's derive a simple random variable $Z_i$ to represent the revenue generated from each sale. Consider the random variable $Z_i$ which is a transformation of our previous outcome variable $Y_i$:

$$Z_i = (1 + \frac{x_{markup}}{100})x_{cost} Y_i$$

A straightforward application of expectations shows that

$$\mathbb E(Z_i) = (1 + \frac{x_{markup}}{100})x_{cost} \mathbb E(Y_i) =  (1 + \frac{x_{markup}}{100})x_{cost} \pi_i$$

where $E(Y_i) = \pi_i$ is a known property of the Bernoulli distribution.

From our logistic regression model, we have a (maximum-likelihood) estimate of $\pi_i$ that we will denote as $\hat \pi_i$. Thus our (maximum-likelihood) estimate of $E(Z_i)$ is

$\widehat{\mathbb E(Z_i)} = (1 + \frac{x_{markup}}{100})x_{cost} \hat \pi_i$.

For the more statistically rigorous, the expectations here are conditional on the covariates $\pmb x_i$ in the logistic regression model and $\widehat{\mathbb E(Z_i)}$ is a maximum-likelihood estimator of $\mathbb E(Z_i)$.

# Appendix 2: Deriving Existence of a Maximum Value

I will warn you that this is a math heavy section.

To determine the existence of an maximum value is a straightforward calculus problem. We are specifically looking for a change in the derivative from positive to negative for values of $x_{markup} > 0$. The tldr; version of this is that $\beta_{markup} < 0$ is the only scenario where a maximum exists and is relatively easy to check.

We need to fully expand the revenue function earlier as

$$
f(x) = c_1(1 + \frac{x}{100})(1 + \exp[-(c_2 + x * \beta )])^{-1}
$$
where I shorten the notation such that $x = x_markup$, $c_1 = x_cost$, and $c_2$ is the rest of the covariates, i.e. $c_2 = -(\beta_0 + x_{cost} * \beta_{cost} + x_{type=shoes} * \beta_{type=shoes} + x_{type = clothes}*\beta_{type = clothes})$. These values are constants in the function as we are only optimizing markup.

Now we calculate the derivative of this function with respect to $x_{markup}$ to be:

$$
f'(x) = c_1 \left [ \left ( \frac{1}{100} \right ) \frac{1}{1 + \exp[-(c_2 + x \beta )]} + \beta (1 + \frac{x}{100}) \left ( \frac{1}{1 + \exp[-(c_2 + x * \beta )]} \right ) \left ( \frac{\exp[-(c_2 + x * \beta )]}{1 + \exp[-(c_2 + x * \beta )} \right ) \right]
$$

Recall that $\pi_i = \frac{1}{1 + \exp[-(c_2 + x \beta )]}$ and therefore $1 - \pi_i = \frac{\exp[-(c_2 + x \beta )]}{1 + \exp[-(c_2 + x \beta )]}$. Rewriting the above derivative gives the more compact equation:

$$
f'(x) = c_1 \left [ \left ( \frac{1}{100} \right ) \pi_i + \beta (1 + \frac{x}{100}) \pi_i (1 - \pi_i) \right]
$$


## Scenario 1: when $\beta < 0$

If $\beta < 0$, then $\displaystyle \lim_{x \to \infty} \pi_i = 0$. Applying this to the derivative reveals that 

$$ 
\displaystyle \lim_{x \to \infty} f'(x) = 0
$$

Suppose that $f'(0) > 0$, then by Rolle's Theorem there exists some value of $x > 0$ where $f'(x) < 0$. Therefore, we only need to check if $f'(0) > 0$ when $\beta < 0$ to know that a maximum exists.

## Scenario 2: When $\beta > 0$

If $\beta > 0$, then $\displaystyle \lim_{x \to \infty} \pi_i = 1$ and applying this to the original function now reveals that

$$
\displaystyle \lim_{x \to \infty} f(x) = \infty
$$

Therefore, no absolute maximum exists in this scenario. To interpret this scenario, the customers are more likely to buy the products the more markup percentage there is, so it would be wise to set the markup to infinity.

## Scenario 3: When $\beta = 0$

For completeness, when $\beta = 0$ then $\pi_i$ is constant with respect to the markup. Therefore the limit for the revenue function is again:

$$
\displaystyle \lim_{x \to \infty} f(x) = \infty
$$

Therefore, no absolute maximum exists in this scenario. Unlike when $\beta > 0$, customers are indifferent to increases in markup so one might as well set the markup to infinity.


# Session Info {.appendix}

Below is my R session info that you may refer to in the event the code is not reproducible. I suggest running all chunks in the order presented and only run them once to ensure you get a reproducible result of this code due to the simulated parts of the analysis.

```{r}
sessionInfo()
```
